The Economic and Health Impact of U.S. Weather Events (1950-2011)
=================================================================

# Synopsis

The data for this analysis was obtained from the U.S. National Oceanic and
Atmospheric Administration's (NOAA) storm database.

# Data Processing

## Obtain and load the data

Download and unzip the data from the NOAA storm database, if this hasn't been
done already:

```{r downloadData}
if(!file.exists('StormData.csv.bz2')) {
    download.file(
        'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
        destfile = 'StormData.csv.bz2', method = 'curl')
}
```

Once the file has been obtained, load the data for analysis (`read.csv` 
automatically decompresses the file, which is very handy):

```{r loadData, cache=TRUE}
data <- read.csv('StormData.csv.bz2')
```

## Isolate relevant data

This study is concerned with the impact of weather events on health and the
economy. The following sections identify, isolate, and process the relevant data
in order to draw sensible conclusions.

An explanation of the data, including the columns identified below, can be found
[here]
(https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

### Event Types

Before investigating health and economic impacts specifically, the different
weather event types (found in the EVTYPE column in the data set) need to be
examined.

```{r eventTypes}
eventTypes <- levels(data$EVTYPE)
```

There are a total of `r length(levels(data$EVTYPE))` different events types. A
cursory investigation, however, reveals a few problems:

#### Some event types are unknown

For example, the very first entry in the `eventTypes` list:

```{r unknownEvent}
eventTypes[1]
```

These are removed:

```{r removeUnknownEvents}
data <- data[data$EVTYPE != eventTypes[1],]
```

#### Some events don't describe events at all

These *summary* events do not make sense in this context:

```{r summaryEvents}
eventTypes[677:684]
```

These are removed:

```{r removeSummaryEvents}
data <- data[-grep('summary', data$EVTYPE, ignore.case=TRUE),]
```

Likewise,

```{r hypothermia}
eventTypes[408:410]
```

Hypothermia may occur as a result of a weather event, but is not a weather
event. These are removed:

```{r removeHypothermiaEvents}
data <- data[-grep('hyperthermia|hypothermia', data$EVTYPE, ignore.case=TRUE),]
```

#### Mixed casing

Entries like these are effectively duplicates, but won't be recognized as such
because of the mixed capital and lower case letters:

```{r mixedCasing}
eventTypes[185:188]
```

Change all EVTYPEs to upper case, strip surrounding whitespace, and correct
unconventional spacing:

```{r toUpperCase}
data$EVTYPE <- toupper(data$EVTYPE)
data$EVTYPE <- gsub('^\\s+|\\s+$', '', data$EVTYPE)
data$EVTYPE <- gsub('\\s+', ' ', data$EVTYPE)
data$EVTYPE <- factor(data$EVTYPE)
```

Now, drop the unused levels:

```{r dropUnusedLevelsPrelim}
d <- droplevels(data$EVTYPE)
eventTypes <- levels(data$EVTYPE)
```

With this preliminary cleanup, there are `r length(eventTypes)` event types
remaining.

#### There is a lot of duplication

For example, the following is a subset of the eventTypes pertaining to some
different kinds of thunderstorms:

```{r freezingPercipitation}
eventTypes[597:606]
```

In light of this messy event classification, the remaining
`r length(eventTypes)` event types will be matched and categorized according to
the most significant terms comprising the events named in the *Storm Data Event
Table* in section 2.1.1 of [Storm Data Preparation]
(https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
. This classification is obtained as follows:

```{r classification}
categories <- list('Astronomical Low Tide' = eventTypes[grep('Astronomical', eventTypes, ignore.case=TRUE)],
                 'Avalanche' = eventTypes[grep('Avalanche', eventTypes, ignore.case=TRUE)],
                 'Blizzard' = eventTypes[grep('Blizzard', eventTypes, ignore.case=TRUE)],
                 'Coastal Flood' = eventTypes[grep('Coastal|Beach', eventTypes, ignore.case=TRUE)],
                 'Cold/Wind Chill' = eventTypes[grep('Wind Chill', eventTypes, ignore.case=TRUE)],
                 'Debris Flow' =  eventTypes[grepl('(?=.*Debris)(?=.*Flow)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Dense Fog' = eventTypes[grepl('(?=.*Dense)(?=.*Fog)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Dense Smoke' = eventTypes[grepl('(?=.*Dense)(?=.*Smoke)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Drought' = eventTypes[grep('Dry|Drought|Below Normal Precipitation', eventTypes, ignore.case=TRUE)],
                 'Dust Devil' = eventTypes[grepl('(?=.*Dust)(?=.*Devil)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Dust Storm' = eventTypes[grepl('(?=.*Dust)(?=.*Storm)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Excessive Heat' = eventTypes[grep('Excessive Heat', eventTypes, ignore.case=TRUE)],
                 'Extreme Cold/Wind Chill' = eventTypes[grepl('(?=.*Bitter)(?=.*Wind)|(?=.*Extreme)(?=.*Wind)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Flash Flood' = eventTypes[grepl('(?=.*Flash)(?=.*Storm)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Flood' = eventTypes[grep('Flood', eventTypes, ignore.case=TRUE)],
                 'Frost/Freeze' = eventTypes[grep('Frost|Freeze', eventTypes, ignore.case=TRUE)],
                 'Funnel Cloud' = eventTypes[grepl('(?=.*Funnel)(?=.*Cloud)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Freezing Fog' = eventTypes[grepl('(?=.*Freezing)(?=.*Fog)', eventTypes, perl=TRUE, ignore.case=TRUE)], 
                 'Hail' = eventTypes[grep('Hail', eventTypes, ignore.case=TRUE)], 
                 'Heat' = eventTypes[grep('Abnormal Warmth', eventTypes, ignore.case=TRUE)], 
                 'Heavy Rain' = eventTypes[grep('Abnormally Wet|Heavy Rain', eventTypes, ignore.case=TRUE)],
                 'Heavy Snow' = eventTypes[grep('Heavy snow', eventTypes, ignore.case=TRUE)], 
                 'High Surf' = eventTypes[grep('High Surf', eventTypes, ignore.case=TRUE)], 
                 'High Wind' = eventTypes[grep('High Wind', eventTypes, ignore.case=TRUE)],  
                 'Hurricane (Typhoon)' = eventTypes[grep('hurricane|typhoon', eventTypes, ignore.case=TRUE)],   
                 'Ice Storm' = eventTypes[grep('Ice Storm', eventTypes, ignore.case=TRUE)],   
                 'Lake-Effect Snow' = eventTypes[grepl('(?=.*Lake)(?=.*Snow)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Lakeshore Flood' = eventTypes[grepl('(?=.*Lake)(?=.*Flood)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Lightning' = eventTypes[grep('Lightning', eventTypes, ignore.case=TRUE)],   
                 'Marine Hail' = eventTypes[grepl('(?=.*Marine)(?=.*Hail)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Marine High Wind' = eventTypes[grepl('(?=.*Marine)(?=.*High)(?=.*Wind)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Marine Strong Wind' = eventTypes[grepl('(?=.*Marine)(?=.*Strong)(?=.*Wind)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Marine Thunderstorm Wind' = eventTypes[grepl('(?=.*Marine)(?=.*Thunderstorm)(?=.*Wind)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Rip Current' = eventTypes[grep('Rip Current', eventTypes, ignore.case=TRUE)], 
                 'Seiche' = eventTypes[grep('Seiche', eventTypes, ignore.case=TRUE)],  
                 'Sleet' = eventTypes[grep('Sleet', eventTypes, ignore.case=TRUE)],  
                 'Storm Surge/Tide' = eventTypes[grep('Blow-out Tide|Coastalstorm|Coastal Storm', eventTypes, ignore.case=TRUE)],
                 'Strong Wind' = eventTypes[grep('Strong Wind', eventTypes, ignore.case=TRUE)],  
                 'Thunderstorm Wind' = eventTypes[grep('Tstm|Thunderstorm', eventTypes, ignore.case=TRUE)],  
                 'Tornado' = eventTypes[grep('Tornado', eventTypes, ignore.case=TRUE)],  
                 'Tropical Depression' = eventTypes[grepl('(?=.*Tropical)(?=.*Depression)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Tropical Storm' = eventTypes[grepl('(?=.*Tropical)(?=.*Storm)', eventTypes, perl=TRUE, ignore.case=TRUE)],
                 'Tsunami' = eventTypes[grep('Tsunami', eventTypes, ignore.case=TRUE)],  
                 'Volcanic Ash' = eventTypes[grep('Volcanic', eventTypes, ignore.case=TRUE)],  
                 'Waterspout' = eventTypes[grep('Waterspout|Water spout', eventTypes, ignore.case=TRUE)],  
                 'Wildfire' = eventTypes[grep('Fire', eventTypes, ignore.case=TRUE)],  
                 'Winter Storm' = eventTypes[grep('Blowing Snow|Snow Squalls', eventTypes, ignore.case=TRUE)],   
                 'Winter Weather' = eventTypes[grep('Black Ice|Accumulated Snow', eventTypes, ignore.case=TRUE)]
            )
```


Apply the new event categories to the data set:

```{r applyNewCategories}
data$EventType <- factor(data$EVTYPE)
levels(data$EventType) <- categories
```

### Health

The following columns contained in the loaded data are relevant here:

1. FATALITIES - The number of direct and indirect fatalities reported
2. INJURIES - The number of direct and indirect injuries reported

Total the number of injuries and fatalities for each `EventType`:

```{r healthImpact}
healthImpact <- by(data, data$EventType, function(x) colSums(x[,c('INJURIES', 'FATALITIES')]))
```

### Economy

The following columns contained in the loaded data are relevant here:

1. PROPDMG - The dollar amount of the property damage
2. PROPDMGEXP - The magnitude of the dollar amount (K = thousands,
M = millions, B = billions)
3. CROPDMG - The dollar amount of damage done to crops
4. CROPDMGEXP - The magnitude of the dollar amount (see PROPDMGEXP) 


# Results


